TEMPERATURES AND METALLICITIES FOR 10,000 M DWARFS\\ IN THE \apogee\ SURVEY

### 2019-03-13 paper/draft.tex
restructure

\begin{abstract}
\thecannon\ is a flexible, data-driven spectral-modeling and parameter-inference framework, demonstrated on high-resolution, near-infrared \sdssiv\ \apogee\ spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H], and detailed abundances) to precisions higher than the \apogee\ survey model-grid pipelines.
The lack of reliable atmospheric models in the near infrared for temperatures cooler than $\sim3500$\,K motivates the extension of this approach to M-dwarf stars, but with calibrated training labels from optical surveys that overlap \apogee.
Using a training sample of 87 M dwarfs in the temperature range $2860 < T_{\rm eff} < 4130$\,K, and metallicty range $-0.5 < \mathrm{[Fe/H]}< 0.5$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex respectively.
The model is accurate in the data space: it accurately predicts \apogee\ spectral pixels at the one-percent level.
We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging M0 to M9, to predictive accuracy of 0.9 types. From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we select a test sample and employ our trained models to determine temperatures, metallicities and spectral subtypes for a conservatively selected sample of M dwarfs in \apogee\ (10,311 sources), and compare our \cannon\ derived \teff\ against color--temperature relations and [Fe/H] against isochrone models in \gaia\ color-magnitude space. Finally we use empirically callibrated relations to estimate radii from $M_K$ absolute magnitudes and \cannon\ temperatures.
This project has implications for the NASA \tess\ Mission and other surveys that involve large samples of M dwarfs. 
\end{abstract}

@@

Low-mass stars (with masses $M_{*} < 0.7\,$\msun and atmospheric temperatures $T_{\rm eff} < 4000\,$K) are by far the most ubiquitous type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and XX of the population by mass. With nuclear fusion timescales of $\tau>10^{12}$ yr \citep{Laughlin:1997}, chemical compositions of M dwarf populations tracing the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, are a unique probe for piecing together galactic structure and evolution \citep{Bochanski:2010}.

@@

Precise measurements of fundamental stellar parameters (including temperatures, masses, radii, and abundances) are key to studying M dwarfs on both galactic and planetary scales. Methods for determining these parameters... temperatures (models, photometry), masses (dynamical)

@@

@@

The data-driven approach of \thecannon\ is ideal in certain cases:
if stellar labels are only known for a small number of stars
but there are spectra taken for many more;
if it is computationally expensive to obtain labels for a star, but
there are many stars that need labels;
if there are spectral models that work in one wavelength range or resolution but not in another.
The context of this paper is the last:
There are working and reliable models of M dwarf stars in the optical, which have been
used successfully, but (as we note above) there isn't as much success with such modeling in the infrared, and not at \apogee\ resolution.

In particular, parameters such as metallicity can be difficult to determine in M dwarfs, as overlapping lines in a spectrum and lack of continuum make it difficult to define features and extract their line strengths. A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07dex) between the adopted metallicities of the FGK primary (under the assumption kinematically similar pairs are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g. \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
However these metallicity callibrations are inaccesible for analysis with \apogee, as the narrow infrared wavelength coverage does not include the relevant callibration indices used by existing studies (such as Ca I or KI lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis. \thecannon\ bridges this gap: it doesn't require using specific lines; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.

@@

\apogee\ is primarily targeted for bright stellar populations particularly red giants, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013} with the objective of studying galactic composition and evolution. However numerous cool, main sequence sources have also been observed either as targets proposed by M dwarf ancilliary surveys, or serendipitously. We determine in this study at least 10,000 M dwarfs have been observed to date.

>>>>

\begin{abstract}
We present a catalog of spectroscopic temperatures, metallicities, spectral types for 10,311 M dwarfs in the \apogee\ survey using \thecannon: a flexible, data-driven spectral-modeling and parameter-inference framework, demonstrated on high-resolution, near-infrared \sdssiv\ \apogee\ spectra of giant stars to estimate stellar labels (\teff, logg, [Fe/H], and detailed abundances) to precisions higher than the \apogee\ survey model-grid pipelines.
Using a training sample of 87 M dwarfs in the temperature range $2860 < T_{\rm eff} < 4130$\,K, and metallicty range $-0.5 < \mathrm{[Fe/H]}< 0.5$\,dex, we train a two-parameter model with predictive accuracy (in cross-validation) to 77\,K and 0.09\,dex respectively.
We also train a one-dimensional spectral type model using 51 M dwarfs with known spectral types ranging M0 to M9, to predictive accuracy of 0.9 types. From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we compare our \cannon\ derived \teff\ against color--temperature relations and \cannon\ derived [Fe/H] against isochrone models in \gaia\ color-magnitude space. Finally we use empirically callibrated relations to estimate masses radii from $M_K$ absolute magnitudes and \cannon\ temperatures.
This project has implications for the NASA \tess\ Mission and other surveys that involve large samples of M dwarfs. 
\end{abstract}

@@

Low-mass stars (with masses $M_{*} < 0.7\,$\msun and atmospheric temperatures $T_{\rm eff} < 4000\,$K) are by far the most ubiquitous type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010}. With nuclear fusion timescales of $\tau>10^{12}$ yr \citep{Laughlin:1997}, chemical compositions of M dwarf populations tracing the nucleosynthetic processes and interstellar mixing of heavy elements from many generations of shorter-lived, high mass stars, are a unique probe for piecing together galactic structure and evolution \citep{Bochanski:2010}.

@@

Precise measurements of fundamental stellar parameters (including temperatures, masses, radii, abundances and kinematics) are key to studying M dwarfs on both galactic and planetary scales. Methods for determining these parameters... temperatures (models, photometry), masses (dynamical)

@@

In particular, parameters such as metallicity can be difficult to determine in M dwarfs, as overlapping lines in a spectrum and lack of continuum make it difficult to define features and extract their line strengths. A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07dex) between the adopted metallicities of the FGK primary (under the assumption kinematically similar pairs are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g. \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
However these metallicity callibrations are inaccesible for analysis with \apogee, as the narrow infrared wavelength coverage does not include the relevant callibration indices used by existing studies (such as Ca I or KI lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis.

@@

The data-driven approach of \thecannon\ is ideal in certain cases:
if stellar labels are only known for a small number of stars
but there are spectra taken for many more;
if it is computationally expensive to obtain labels for a star, but
there are many stars that need labels;
if there are spectral models that work in one wavelength range or resolution but not in another.
The context of this paper is the last:
There are working and reliable models of M dwarf stars in the optical, which have been
used successfully, but (as we note above) there isn't as much success with such modeling in the infrared, and not at \apogee\ resolution.
\thecannon\ bridges this gap: it doesn't require using specific lines; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.

@@

\apogee\ is primarily targeted for bright stellar populations particularly red giants, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013} with the objective of studying galactic composition and evolution. However numerous cool, main sequence sources have also been observed either as targets proposed by M dwarf ancilliary surveys, or serendipitously. We determine in this study at least 10,000 M dwarfs have been observed by \apogee\ to date.
